Cheminformatics: The Computer Science of Chemical Discovery

### 2011-07-19 paper.tex

\title{Cheminformatics: The Computer Science of Chemical Discovery}
\numberofauthors{9}
\author{
\alignauthor
Joerg Kurt Wegner\\
       \affaddr{Tibotec BVBA}\\
       \affaddr{Turnhoutseweg 30}\\
       \affaddr{2340 Beerse Turnhout, Belgium}\\
       \email{jwegner@its.jnj.com}
% 2nd. author
\alignauthor
Aaron Sterling\\
       \affaddr{Department of Computer Science}\\
       \affaddr{Iowa State University}\\
       \affaddr{Ames, Iowa, USA}\\
       \email{sterling@iastate.edu}
% 3rd author
\alignauthor
Rajarshi Guha\\
\affaddr{NIH Center for Translational Therapeutics}\\
\affaddr{9800 Medical Center Drive}\\
\affaddr{Rockvile, MD 20850}\\
\email{guhar@mail.nih.gov}
}

\additionalauthors{Additional authors:
Andreas Bender (University of Cambridge, email: {\texttt{bender.andreas@gmail.com}}),
Jean-Loup Faulon (University of Evry, email: {\texttt{Jean-Loup.Faulon@issb.genopole.fr}}),
Janna Hastings (European Bioinformatics Institute, Cambridge, UK, email: {\texttt{janna.hastings@gmail.com}}),
Noel O'Boyle (University College, Cork, Ireland, email: {\texttt{baoilleach@gmail.com}}),
John Overington (European Bioinformatics Institute, Cambridge, UK, email: {\texttt{jpo@ebi.ac.uk}}),
Herman Van Vlijmen (Tibotec, Beerse, Belgium, email: {\texttt{hvvlijme@its.jnj.com}}), and
Egon Willighagen (Karolinska Institutet, Stockholm, Sweden, email: {\texttt{egon.willighagen@ki.se}})
.}
\date{25 June 2011}

\maketitle
\begin{abstract}
  One of the most prominent success stories in all of science over the last decade has
  been the advance of bioinformatics: interdisciplinary collaboration between
  computer scientists and molecular biologists led to the sequencing the human genome
  (among many other accomplishments), because researchers connected biological
  concepts to the theory of string algorithms. However, despite this great
  success, few computer scientists are familiar with a related (and older!)
  discipline---cheminformatics, the use of computers to represent
  small molecules structures and analyze their properties based on aspects
  of their structures. Cheminformatics has a wide applicability
  including the discovery of novel drug candidates to optimizing
  certain physicochemical properties of small molecules. Until
  recently, the data and techniques of chemoinformatics have been closely guarded
  secrets of companies whose financial success depended on being the first to produce
  the new ``miracle molecule.'' Only within the last decade---and largely because of
  chemists volunteering their time for an Open Science ``movement''---do researchers
  now have access to freely available software packages and databases of tens of
  millions of chemicals. As a result, chemists now confront a variety of unsolved
  algorithmic problems that could not have been tackled a decade ago, but whose
  solutions are critical to research ranging from determining the behavior of small
  molecules in biological pathways, to finding a cure for malaria.
\end{abstract}

\category{H.4}{Information Systems Applications}{Miscellaneous}
\terms{
\begin{itemize}
\item \textit{Tautomerism} - Tautomers are isomers (same molecular formula but different structural formula) of organic compounds that readily interconvert by a chemical reaction called tautomerization.
\item \textit{Protonation and protonation state} - In chemistry, protonation is the addition of a proton (H$^+$) to an atom, molecule, or ion. The different states (different number of H$^+$ or on different atoms) are called protonation states for a molecular compound (e.g. de-protonated, mono-protonated, bi-protonated, or any variations of those states).
\item \textit{BioAssay} - A system measuring the biological activity of a chemical compound on a particular biological assay (with all its parameters).
\item \textit{Metabolites} - Metabolites are the intermediates and products of metabolism. The term metabolite is usually restricted to small molecules (chemical compounds).
\item \textit{Mass spectrometry} - Mass spectrometry (MS) is an analytical technique that measures the mass-to-charge ratio of charged particles. It is used for determining masses of particles, for determining the elemental composition of a sample or molecule, and for elucidating the chemical structures of molecules, such as peptides and other chemical compounds.
\item \textit{Ligand (coordination chemistry)} - an atom, ion, or functional group that donates one or more of its electrons through a coordinate covalent bond to one or more central atoms or ions.
\item \textit{Ligand (biochemistry)} - a substance that binds to a protein.
\item \textit{Receptor (biochemistry)} - in biochemistry, a protein molecule that receives and responds to a neurotransmitter, or other substance.
\item \textit{Proteochemometric modeling} - combining data mining for chemical compounds (ligands) and proteins within the same machine learning framework.
\item \textit{BioDegradation} - is the chemical dissolution of materials by bacteria or other biological means.
\item \textit{Reaction catalyses} - Catalysis is the change in rate of a chemical reaction due to the participation of a substance called a catalyst.
\item \textit{Conformation and conformational isomerism} - In chemistry, conformational isomerism is a form of stereoisomerism in which the isomers can be interconverted exclusively by rotations about formally single bonds. Single possible states of compounds are called conformers.
\item \textit{SMARTS} - SMiles ARbitrary Target Specification is a language for specifying substructural patterns. in molecules
\end{itemize}
}
\keywords{cheminformatics, graphs, chemistry, databases, knowledge management, open source, open data}

@@

Most cheminformatics applications rely on large databases of chemical structures
and their properties. Organising and maintaining such databases, searching based
on features of chemical structures, and clustering similar structures together,
are essential utilities to enable many scientific applications. Yet, each one of
those challenges poses its unique computer science complexities.

There is a considerable progress in open chemical and bioactivity databases like
ChEMBL and NCBI-PubChem, which are both freely available. Still, the integration
of chemical databases with one other remains a challenge, not only due to the
normalization problems in the chemical and bioactivity landscape, but also due
to sheer data volume. As mentioned earlier, we have to be able to
search in complex data, for example multi-labelled molecular graphs, where
graph labels can be different. Still, we need to be able to capture
molecular graphs in a risk-bounded machine encoded way.

Chemical graphs can differ based on the manner in which the structures
are drawn. Also chemical behaviors can result in multiple encodings of
``the same'' molecule and its tautomeric states, different protonation
states, or many variations of implicit chemical expert
knowledge. Checking whether two graphs represent the same chemical is
an expensive operation (the next section will discuss this in-depth);
thus, a representation of the structure that is unique and invariant
with respect to how the structure is drawn is needed for
identification. An early effort in this direction was the Simplified
Molecular Input Line Specification (SMILES). However, several
different SMILES implementations became available, with no universal
standard of interpretation. After years of research and various
commercial products, there is today no single public standard for a
line-notation of molecular labeled graphs, which is available to a
broad audience (academic and
commercial). %The very same problem does also exist for
%bioassay ontologies and the lack of options to normalize the property spaces of
%chemical compounds, e.g. biological assays or clinical data outcome
%(biobanking).

@@

InChI or other identity-mapping algorithms allow for exact searching (within the
limitations of the algorithms), but two other requirements for scientific
discovery based on chemical databases are \emph{substructure searching} and \emph{similarity
searching}. In substructure searching, the database is searched for a specified
wholly contained part of the structures in the database; in similarity
searching, the database is searched for structures similar to a provided search
structure. Chemical search packages are often implemented and optimised for a
given database technology, for example, the OrChem package is an open source
chemical search package for the Oracle database \cite{rijnbeek2009}.

Graph substructure matching is a variant of \emph{graph isomorphism detection},which is
well known to be computationally intractable [3]. To execute a graph isomorphism
search across a full chemical database of thousands or even millions of
structures is completely infeasible. For that reason, complex heuristics must be
used to drastically filter the candidate search space. Structure-based
\emph{fingerprints} are commonly used for this purpose. Fingerprints encode
characteristic features of a given chemical structure in a boolean array, or
bitmap. The fingerprint is created by an algorithm which generates bitmap
patterns for each atom in the structure; then each atom and its nearest
neighbours, including the bonds between them; then each group of atoms connected
by paths of increasing lengths up to the maximum implemented for the algorithm
(commonly around eight). The bitmap patterns are added together into the final
fingerprint. If all bits in a query fingerprint are also present in the target
fingerprint of a stored database structure, only then is the structure subjected
to the computationally expensive subgraph matching algorithm. Bit operations are
very fast and independent of the number of atoms in a structure due to the fixed
length of the fingerprint.

Fingerprints also facilitate the rapid calculation of quantitative
structure-based similarity measures. An example measure of similarity is the
\emph{Tanimoto coefficient}, calculated as the ratio $T(a,b) = \frac{c}{(a+b−c)}$,
where c is the count of bits on in the same position in both the two
fingerprints, a is the count of bits on in the first structure, and b is the
count of bits on in the second structure. The Tanimoto coefficient varies in the
range 0.0 -- 1.0, with a score of 1.0 indicating that the two structures are very
similar (i.e. their fingerprints are the same). Structure-based similarity is of
limited use in grouping of chemicals based on non-structural features, such as
shared bioactivity profiles. For this purpose, a large-scale and flexible
classification structure is needed for chemicals and their linked biological
information. One such answer to this need is provided in the form of \emph{ontologies}.
An ontology is a formal specification of entities and their relationships in a
domain of interest. It is structured around an underlying logical representation
such as Description Logic ~\cite{baaderdl2007}. The logic-based representation allows
automated reasoning to derive new knowledge (inferences) from the knowledge
encoded, making the representation both compact and powerful.
Ontologies are already in widespread use across the biomedical domain. A
similarity measure based on the information encoded in ontologies is termed
\emph{semantic similarity}, and such measures have already been applied to enhance
chemical classification in several problem areas \cite{couto2010}. Many open
questions still remain about the best way to combine the information encoded in
chemical structures with chemical ontologies \cite{hastingsowled2010}.

@@

The future still leaves many challenges ahead. Using external/public databases
with chemical and bioactivity data remains a challenge due to differences in
ontologies, synchronization and maintenance, and efficient
substructure and similarity search within complex data types, aka compounds. A
collaboration with external parties, e.g. contract research, poses other
problems with duplicate checking for compounds (e.g. tautomeric or salt
structures), efficient data synchronization, most often a double-registration is
required for fulfilling the data quality processes on both sides. If external
partners are small they often do not even have sufficient IT resources
themselves and rely on external services themselves. Cloud services could help
not only to provide a service infrastructure for all parties involved,
but also to provide the required private/public access toolbox. Furthermore
there are still many data sources not being indexed properly, since chemistry
patents are often cryptic and image/text mining remains a puzzle. One could
argue that the image/text mining in scientific journals is less challenging; in
fact, it remains a huge challenge and many chemists do not trust automated
non-expert validated mining sources. The closed CAS (Chemical Abstract Service)
is a highly trusted source, and public chemistry and bioactivity space has to
improve quality and interconnectivity. Efforts like Chembl and PubChem-BioAssay
are on the right track. Still, improving data quality and standards between
public and closed sources will be absolutely critical for ensuring a constant
growth, usage, and collaboration scenarios between private and public parties.
Only when the technical and scientific questions are solved can we ensure that
legal questions about contribution metrics will get enough room to grow.

@@

A number of toolkits are available including the Chemistry Development
Kit (CDK) \cite{steinbeck2003}, OpenBabel, RDKit and Indigo. These toolkits are written in
Java or C++ (though toolkits written in the latter invariably have
bindings to a variety of languages). Such toolkits are used, for
example, to read/write chemical file formats, manipulate chemical
structures, measure similarity of molecules, search for substructures,
and generate 2D diagrams. The underlying algorithms used include graph
algorithms (e.g. maximal common substructure, canonical labelling of
graphs), geometrical methods (e.g. Kabsch alignment) and vector
manipulation (e.g. converting coordinates in various systems, 3D
structure generation). As well as being of direct use to practising
cheminformaticians in their day-to-day work, many applications have
been developed that rely on these toolkits to handle chemical data;
for example, the molecular viewer Avogadro
(http://avogadro.openmolecules.net) uses Open Babel, while the
molecular workbench Bioclipse \cite{Bioclipse2} uses the CDK.

The various toolkits have many features in common and at the same time
have certain distinguishing features. For example, the CDK implements
a large collection of molecular descriptors. On the other hand,
OpenBabel has robust support for 3D structure generation and
optimization and supports the interconversion of a huge number of
chemical file formats.

@@

Given that there is ample scope for software engineering (algorithm
implementation, code quality and analysis, build systems etc), both
the CDK and OpenBabel are open to new contributions. Both are driven
via public, active mailing lists and public issue tracking systems.
Development questions are actively and openly discussed to arrive at a
consensus and contributors are encouraged to get involved with any of
the following areas: development of new features, improvement of
existing features, performance optimisation, fixing bugs,
documentation, website development, reporting bugs, and helping users
on the mailing list. Anyone interested should email the respective
developer's mailing list (either cdk-devel@lists.sf.net or
openbabel-devel@lists.sf.net) with their interests and background, and
a mentor will be assigned to help them join the project.

\section{Conclusions}
\label{sec:conclusions}
Cheminformatics, the computer science of chemical discovery, is an
industrial and academic discipline with roots dating back to the
1960s, with the creation of DENDRAL, the first expert system of any
kind.  Much of its development from the 1960s until the new millenium
was performed in secret, because of competition in the pharmaceutical
industry.  While chemists were among the first scientists to embrace
web-based tools of knowledge management, research into fundamental
algorithmic questions of structure enumeration and database management
(and a host of other topics) have suffered, because of limited
academic access to either tools or data.  As we enter the second
decade of the 21st century, however, this landscape is rapidly
shifting.  There now exist free-of-charge open-source cheminformatics
toolkits, and open databases with tens of millions of compounds.  We
believe the prospects for academic and industrial collaboration
between chemists and computer scientists are quite bright, and we
would like to encourage more computer scientists to participate in
cheminformatics research.  There is an admittedly steep learning
curve, as the structural issues faced by chemists are often too
detailed and subtle to admit as clean an abstraction as ``algorithms
on a string,'' the way many bioinformatics can be abstracted into a CS
framework.  Nevertheless, because of the changing landscape, there are
many research questions we are only now in position to try to answer:
for a theorist, graph-theoretic questions of three-dimensional
enumeration; for a database designer, more effective search algorithms
and ontologies; for a practical programmer, expansion of the many
open-source cheminformatics projects.  If more chemists think
algorithmically, and more computer scientists think ``chemically,''
%we may discover ``miracle drugs'' more consistently than ever before.
we are much better positioned taking not only ``simple'' chemicals
into account (only one identifier and isomer possible per molecule),
but also their complex relationships, transformations, and
combinatorial challenges into account. This will allow us to
normalize, search, and use all chemical information, around us and
within us.

>>>>

\title{Cheminformatics: The Computer Science of Chemical Discovery}
\numberofauthors{9}
\author{
\alignauthor
Joerg Kurt Wegner\\
       \affaddr{Tibotec BVBA}\\
       \affaddr{Turnhoutseweg 30}\\
       \affaddr{2340 Beerse Turnhout, Belgium}\\
       \email{jwegner@its.jnj.com}
% 2nd. author
\alignauthor
Aaron Sterling\\
       \affaddr{Department of Computer Science}\\
       \affaddr{Iowa State University}\\
       \affaddr{Ames, Iowa, USA}\\
       \email{sterling@iastate.edu}
% 3rd author
\alignauthor
Rajarshi Guha\\
\affaddr{NIH Center for Translational Therapeutics}\\
\affaddr{9800 Medical Center Drive}\\
\affaddr{Rockville, MD 20850}\\
\email{guhar@mail.nih.gov}
}

\additionalauthors{Additional authors:
Andreas Bender (University of Cambridge, email: {\texttt{bender.andreas@gmail.com}}),
Jean-Loup Faulon (University of Evry, email: {\texttt{Jean-Loup.Faulon@issb.genopole.fr}}),
Janna Hastings (European Bioinformatics Institute, Cambridge, UK, email: {\texttt{janna.hastings@gmail.com}}),
Noel O'Boyle (University College Cork, Cork, Ireland, email: {\texttt{baoilleach@gmail.com}}),
John Overington (European Bioinformatics Institute, Cambridge, UK, email: {\texttt{jpo@ebi.ac.uk}}),
Herman Van Vlijmen (Tibotec, Beerse, Belgium, email: {\texttt{hvvlijme@its.jnj.com}}), and
Egon Willighagen (Karolinska Institutet, Stockholm, Sweden, email: {\texttt{egon.willighagen@ki.se}})
.}
\date{25 June 2011}

\maketitle
\begin{abstract}
  One of the most prominent success stories in all of science over the last decade has
  been the advance of bioinformatics: interdisciplinary collaboration between
  computer scientists and molecular biologists led to the sequencing the human genome
  (among many other accomplishments), because researchers connected biological
  concepts to the theory of string algorithms. However, despite this great
  success, few computer scientists are familiar with a related (and older!)
  discipline---cheminformatics, the use of computers to represent
  small molecules structures and analyze their properties based on aspects
  of their structures. Cheminformatics has a wide applicability
  from the discovery of novel drug candidates to optimizing
  certain physicochemical properties of small molecules. Until
  recently, the data and techniques of chemoinformatics have been closely guarded
  secrets of companies whose financial success depended on being the first to produce
  the new ``miracle molecule.'' Only within the last decade---and largely because of
  chemists volunteering their time for an Open Science ``movement''---do researchers
  now have access to freely available software packages and databases of tens of
  millions of chemicals. As a result, chemists now confront a variety of unsolved
  algorithmic problems that could not have been tackled a decade ago, but whose
  solutions are critical to research ranging from determining the behavior of small
  molecules in biological pathways, to finding a cure for malaria.
\end{abstract}

\category{H.4}{Information Systems Applications}{Miscellaneous}
\terms{
\begin{itemize}
\item \textit{Tautomerism} - Tautomers are isomers (same molecular formula but different structural formula) of organic compounds that readily interconvert by a chemical reaction called tautomerization.
\item \textit{Protonation and protonation state} - In chemistry, protonation is the addition of a proton (H$^+$) to an atom, molecule, or ion. The different states (different number of H$^+$ or on different atoms) are called protonation states for a molecular compound (e.g. de-protonated, mono-protonated, di-protonated, or any variations of those states).
\item \textit{Bioassay} - A system measuring the biological activity of a chemical compound on a particular biological assay (with all its parameters).
\item \textit{Metabolites} - Metabolites are the intermediates and products of metabolism. The term metabolite is usually restricted to small molecules (chemical compounds).
\item \textit{Mass spectrometry} - Mass spectrometry (MS) is an analytical technique that measures the mass-to-charge ratio of charged particles. It is used for determining masses of particles, for determining the elemental composition of a sample or molecule, and for elucidating the chemical structures of molecules, such as peptides and other chemical compounds.
\item \textit{Ligand (coordination chemistry)} - an atom, ion, or functional group that donates one or more of its electrons through a coordinate covalent bond to one or more central atoms or ions.
\item \textit{Ligand (biochemistry)} - a substance that binds to a protein.
\item \textit{Receptor (biochemistry)} - in biochemistry, a protein molecule that receives and responds to a neurotransmitter, or other substance.
\item \textit{Proteochemometric modeling} - combining data mining for chemical compounds (ligands) and proteins within the same machine learning framework.
\item \textit{Biodegradation} - is the chemical dissolution of materials by bacteria or other biological means.
\item \textit{Reaction catalysis} - Catalysis is the change in rate of a chemical reaction due to the participation of a substance called a catalyst.
\item \textit{Conformation and conformational isomerism} - In chemistry, conformational isomerism is a form of stereoisomerism in which the isomers can be interconverted exclusively by rotations about formally single bonds. Single possible states of compounds are called conformers.
\item \textit{SMARTS} - SMiles ARbitrary Target Specification (SMARTS) is a language for specifying substructure patterns in molecules
\end{itemize}
}
\keywords{cheminformatics, graphs, chemistry, databases, knowledge management, open source, open data}

@@

Most cheminformatics applications rely on large databases of chemical structures
and their properties. Organizing and maintaining such databases, searching based
on features of chemical structures, and clustering similar structures together,
are essential utilities to enable many scientific applications. Yet, each one of
those areas poses its unique computer science challenges.

There is a considerable progress in open chemical and bioactivity databases like
ChEMBL and NCBI-PubChem, which are both freely available. Still, the integration
of chemical databases with each other remains a challenge, not only due to the
normalization problems in the chemical and bioactivity landscape, but also due
to sheer data volume. As mentioned earlier, we have to be able to
search in complex data, for example multi-labelled molecular graphs, where
graph labels can be different. Still, we need to be able to capture
molecular graphs in a risk-bounded machine encoded way.

Chemical graphs can differ based on the manner in which the structures
are drawn. Also chemical behaviors can result in multiple encodings of
``the same'' molecule and its tautomeric states, different protonation
states, or many variations of implicit chemical expert
knowledge. Checking whether two graphs represent the same chemical is
an expensive operation (the next section will discuss this in-depth);
thus, a representation of the structure that is unique and invariant
with respect to how the structure is drawn is needed for
identification. An early effort in this direction was the Simplified
Molecular Input Line Specification (SMILES), and specifically the
generation of a single canonical SMILES for a particular structure. However,
with time several
different SMILES implementations became available with no universal
standard of generating a canonical SMILES. After years of research and various
commercial products, there is today no single public standard for a
line-notation of molecular labeled graphs, which is available to a
broad audience (academic and
commercial). %The very same problem does also exist for
%bioassay ontologies and the lack of options to normalize the property spaces of
%chemical compounds, e.g. biological assays or clinical data outcome
%(biobanking).

@@

InChI or other identity-mapping algorithms allow for exact searching (within the
limitations of the algorithms), but two other requirements for scientific
discovery based on chemical databases are \emph{substructure searching} and \emph{similarity
searching}. In substructure searching, the database is searched for a specified
wholly contained part of the structures in the database; in similarity
searching, the database is searched for structures similar to a provided search
structure. Chemical search packages are often implemented and optimized for a
given database technology, for example, the OrChem package is an open source
chemical search package for the Oracle database \cite{rijnbeek2009}.

Graph substructure matching is a variant of \emph{graph isomorphism detection},which is
well known to be computationally intractable [3]. To execute a graph isomorphism
search across a full chemical database of thousands or even millions of
structures is completely infeasible. For that reason, complex heuristics must be
used to drastically filter the candidate search space. Structure-based
\emph{fingerprints} are commonly used for this purpose. Fingerprints encode
characteristic features of a given chemical structure in a Boolean array, or
bitmap. The fingerprint is created by an algorithm which generates bitmap
patterns for each atom in the structure; then each atom and its nearest
neighbors, including the bonds between them; then each group of atoms connected
by paths of increasing lengths up to the maximum implemented for the algorithm
(commonly around eight). The bitmap patterns are added together into the final
fingerprint. If all bits in a query fingerprint are also present in the target
fingerprint of a stored database structure, only then is the structure subjected
to the computationally expensive subgraph matching algorithm. Bit operations are
very fast and independent of the number of atoms in a structure due to the fixed
length of the fingerprint.

Fingerprints also facilitate the rapid calculation of quantitative
structure-based similarity measures. An example measure of similarity is the
\emph{Tanimoto coefficient}, calculated as the ratio $T(a,b) = \frac{c}{(a+b-c)}$,
where c is the count of bits on in the same position in both the two
fingerprints, a is the count of bits on in the first structure, and b is the
count of bits on in the second structure. The Tanimoto coefficient varies in the
range 0.0 -- 1.0, with a score of 1.0 indicating that the two structures are very
similar (i.e. their fingerprints are the same). Structure-based similarity is of
limited use in grouping of chemicals based on non-structural features, such as
shared bioactivity profiles. For this purpose, a large-scale and flexible
classification structure is needed for chemicals and their linked biological
information. One such answer to this need is provided in the form of \emph{ontologies}.
An ontology is a formal specification of entities and their relationships in a
domain of interest. It is structured around an underlying logical representation
such as Description Logic ~\cite{baaderdl2007}. The logic-based representation allows
automated reasoning to derive new knowledge (inferences) from the knowledge
encoded, making the representation both compact and powerful.
Ontologies are already in widespread use across the biomedical domain. A
similarity measure based on the information encoded in ontologies is termed
\emph{semantic similarity}, and such measures have already been applied to enhance
chemical classification in several problem areas \cite{couto2010}. Many open
questions still remain about the best way to combine the information encoded in
chemical structures with chemical ontologies \cite{hastingsowled2010}.

@@

The future still leaves many challenges ahead. Using external/public databases
with chemical and bioactivity data remains a challenge due to differences in
ontologies, synchronization and maintenance, and efficient
substructure and similarity search within complex data types
(compounds). A
collaboration with external parties, e.g. contract research, poses other
problems with duplicate checking for compounds (e.g. tautomeric or salt
structures), efficient data synchronization, most often a double-registration is
required for fulfilling the data quality processes on both sides. If external
partners are small they often do not even have sufficient IT resources
themselves and rely on external services themselves. Cloud services could help
not only to provide a service infrastructure for all parties involved,
but also to provide the required private/public access toolbox. Furthermore
there are still many data sources not being indexed properly, since chemistry
patents are often cryptic and image/text mining remains a puzzle. One could
argue that the image/text mining in scientific journals is less challenging; in
fact, it remains a huge challenge and many chemists do not trust automated
non-expert validated mining sources. The closed CAS (Chemical Abstract Service)
is a highly trusted source, and public chemistry and bioactivity space has to
improve quality and interconnectivity. Efforts like ChEMBL and PubChem-BioAssay
are on the right track. Still, improving data quality and standards between
public and closed sources will be absolutely critical for ensuring a constant
growth, usage, and collaboration scenarios between private and public parties.
Only when the technical and scientific questions are solved can we ensure that
legal questions about contribution metrics will get enough room to grow.

@@

The extent of open-source software in cheminformatics has somewhat
lagged behind related fields such as bioinformatics, partly
due to the existence of a lucrative pharmaceutical market for
proprietary software. However, in recent years there has been a
rise in the amount of open-source cheminformatics software. One
contribution to this has been the Blue Obelisk group
(http://blueobelisk.org) of chemists, who have worked together to
create interoperable open-source chemistry software (among other
activities). Now there are a number of open-source toolkits available
including the Chemistry Development
Kit (CDK) \cite{steinbeck2003}, Open Babel, RDKit and Indigo. These toolkits are written in
Java or C++ (though toolkits written in the latter invariably have
bindings to a variety of languages). Such toolkits are used, for
example, to read/write chemical file formats, manipulate chemical
structures, measure similarity of molecules, search for substructures,
and generate 2D diagrams. The underlying algorithms used include graph
algorithms (e.g. maximal common substructure, canonical labelling of
graphs), geometrical methods (e.g. Kabsch alignment) and vector
manipulation (e.g. converting coordinates in various systems, 3D
structure generation). As well as being of direct use to practising
cheminformaticians in their day-to-day work, many applications have
been developed that rely on these toolkits to handle chemical data;
for example, the molecular viewer Avogadro
(http://avogadro.openmolecules.net) uses Open Babel, while the
molecular workbench Bioclipse \cite{Bioclipse2} uses the CDK.

The various toolkits have many features in common and at the same time
have certain distinguishing features. For example, the CDK implements
a large collection of molecular descriptors. On the other hand,
Open Babel has robust support for 3D structure generation and
optimization and supports the interconversion of a huge number of
chemical file formats.

@@

Given that there is ample scope for software engineering (algorithm
implementation, code quality and analysis, build systems etc), both
the CDK and Open Babel are open to new contributions. Both are driven
via public, active mailing lists and public issue tracking systems.
Development questions are actively and openly discussed to arrive at a
consensus and contributors are encouraged to get involved with any of
the following areas: development of new features, improvement of
existing features, performance optimization, fixing bugs,
documentation, website development, reporting bugs, and helping users
on the mailing list. Anyone interested should email the respective
developer's mailing list (either cdk-devel@lists.sf.net or
openbabel-devel@lists.sf.net) with their interests and background, and
a mentor will be assigned to help them join the project.

\section{Conclusions}
\label{sec:conclusions}
Cheminformatics, the computer science of chemical discovery, is an
industrial and academic discipline with roots dating back to the
1960s, with the creation of DENDRAL, the first expert system of any
kind.  Much of its development from the 1960s until the new millenium
was performed in secret, because of competition in the pharmaceutical
industry.  While chemists were among the first scientists to embrace
web-based tools of knowledge management, research into fundamental
algorithmic questions of structure enumeration and database management
(and a host of other topics) have suffered, because of limited
academic access to either tools or data.  As we enter the second
decade of the 21st century, however, this landscape is rapidly
shifting.  There now exist free-of-charge open-source cheminformatics
toolkits, and open databases with tens of millions of compounds.  We
believe the prospects for academic and industrial collaboration
between chemists and computer scientists are quite bright, and we
would like to encourage more computer scientists to participate in
cheminformatics research.  There is an admittedly steep learning
curve, as the structural issues faced by chemists are often too
detailed and subtle to admit as clean an abstraction as ``algorithms
on a string,'' the way many bioinformatics algorithms can be abstracted into a CS
framework.  Nevertheless, because of the changing landscape, there are
many research questions we are only now in position to try to answer:
for a theorist, graph-theoretic questions of three-dimensional
enumeration; for a database designer, more effective search algorithms
and ontologies; for a practical programmer, expansion of the many
open-source cheminformatics projects.  If more chemists think
algorithmically, and more computer scientists think ``chemically,''
%we may discover ``miracle drugs'' more consistently than ever before.
we are much better positioned taking not only ``simple'' chemicals
into account (only one identifier and isomer possible per molecule),
but also their complex relationships, transformations, and
combinatorial challenges into account. This will allow us to
normalize, search, and use all chemical information, around us and
within us.
